The guts of the dispRity package

### 2017-12-05 inst/gitbook/03_specific-tutorials.Rmd
updated manual
updated gitbook

Before each section, make sure you loaded the Beck and Lee (2014) data (see [example data](#example-data) for more details).

@@

 *  `acctran` where the data chosen along the branch is always the one of the descendant
 *  `deltran` where the data chosen along the branch is always the one of the ancestor
 *  `random` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
 *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length
 *  `punctuated` where the data chosen along the branch is both the descendant and the ancestor with an even probability
 *  `gradual` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length

> Note that the four first models are "fixed": the selected data is always either the one of the descendant or the ancestor.
The two last models are "probabilistic": the data from both the descendant and the ancestor is used with an associate probability.
These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.

@@

>>>>

Before each section, make sure you loaded the @beckancient2014 data (see [example data](#example-data) for more details).

@@

 * **Punctuated models**
   *  `acctran` where the data chosen along the branch is always the one of the descendant
   *  `deltran` where the data chosen along the branch is always the one of the ancestor
   *  `random` where the data chosen along the branch is randomly chosen between the descendant or the ancestor
   *  `proximity` where the data chosen along the branch is either the descendant or the ancestor depending on branch length

 * **Gradual models**
   *  `equal.split` where the data chosen along the branch is both the descendant and the ancestor with an even probability
   *  `gradual.split` where the data chosen along the branch is both the descendant and the ancestor with a probability depending on branch length

> Note that the four first models are a proxy for punctuated evolution: the selected data is always either the one of the descendant or the ancestor.
In other words, changes along the branches always occur at either ends of it.
The two last models are a proxy for gradual evolution: the data from both the descendant and the ancestor is used with an associate probability.
These later models perform better when bootstrapped, effectively approximating the "intermediate" state between and the ancestor and the descendants.

@@

### NPMANOVA in `dispRity`

One often useful test to apply to multidimensional data is the permutational multivariate analysis of variance based on distance matrices `vegan::adonis`.

```{r}
library(vegan)
data(dune)
data(dune.env)

## Beta diversity distance matrix
betad <- betadiver(dune, "z")

## Create a pairwise distance matrix
make.pairwise.dist.matrix <- function(matrix) {
    if(class(matrix) == "matrix" || class(matrix) == "data.frame") {
        ## Simply calculate the distances
        return(dist(matrix))
    } else {
        ## Transform the distances into a matrix
        pair_distances <- matrix
        pairwise_dist <- matrix(NA, nrow = nrow(matrix), ncol = nrow(matrix))
        colnames(pairwise_dist) <- rownames(pairwise_dist) <- rownames(matrix)
        diag(pairwise_dist) <- 0
        pairwise_dist[lower.tri(pairwise_dist)] <- pair_distances
        pairwise_dist <- as.dist(pairwise_dist)
    }
}
pairwise_dist <- make.pairwise.dist.matrix(dune)

## Create the factor group
factors <- data.frame(factor = dune.env$Management)

## Testing adonis
test1 <- adonis(betad ~ factor, factors)
test2 <- adonis(pairwise_dist ~ factor, factors)

set.seed(1)
## Check if it works for mammals data set
data(BeckLee_mat50)
data(BeckLee_tree)
## Crown stem group
crown_stem <- crown.stem(BeckLee_tree, inc.nodes = FALSE)
## Dummy group
group_sample <- sample(1:50, 25)
group1 <- rownames(BeckLee_mat50)[group_sample]
group2 <- rownames(BeckLee_mat50)[-group_sample]
dummy_group <- list(group1 = group1, group2 = group2)
## Equal groups
equal_group <- list(groupA = rownames(BeckLee_mat50), groupB = rownames(BeckLee_mat50))

## Create the subsets
subset_crown_stem <- custom.subsets(BeckLee_mat50, group = crown_stem)
subset_dummy_group <- custom.subsets(BeckLee_mat50, group = dummy_group)
subset_equal_group <- custom.subsets(BeckLee_mat50, group = equal_group)

## Calculate disparity
disp_crown_stem <- dispRity(subset_crown_stem, metric = c(median, centroids))
disp_dummy_group <- dispRity(subset_dummy_group, metric = c(median, centroids))
disp_equal_group <- dispRity(subset_equal_group, metric = c(median, centroids))

summary(disp_crown_stem)
summary(disp_dummy_group)
summary(disp_equal_group)

## Function for creating the factors table
make.factors <- function(data) {
    ## Extracting the factors
    factors <- names(sort(unlist(lapply(data$subsets, function(X) return(X$elements)))))
    ## Trimming the factors names
    group_names <- names(data$subsets)
    for(group in 1:length(group_names)) {
        factors[grep(group_names[group], factors)] <- group_names[group]
    }
    return(data.frame(factor = factors, row.names = rownames(data$matrix)))
}

adonis.dispRity <- function(data, ...) {
  ## Create the factor table
  factors <- make.factors(data)

  ## Run adonis
  return(adonis(make.pairwise.dist.matrix(data$matrix)~factor, factors, ...))
}

### 2017-12-05 inst/gitbook/04_making-up-data.Rmd
updated gitbook

The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].

>>>>

The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;@OReilly2017].

### 2017-12-05 inst/gitbook/05_guts.Rmd
updated gitbook

 * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"punctuated"`, `"gradual"`). This element generated only once via `time.subsets()` and `custom.subsets()`.
 * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use. This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
 * `$call$bootstrap`: this is a `list` containing three elements:
    * `[[1]]`: the number of bootstrap replicates (`numeric`)
    * `[[2]]`: the bootstrap method (`character`)
    * `[[3]]`: the rarefaction levels (`numeric` vector)
 * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`. These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).

>>>>

 * `$call$subsets`: a vector of `character` with information on the subsets type (either `"continuous"`, `"discrete"` or `"custom"`) and their eventual model (`"acctran"`, `"deltran"`, `"random"`, `"proximity"`, `"equal.split"`, `"gradual.split"`). This element generated only once via `time.subsets()` and `custom.subsets()`.
 * `$call$dimensions`: either a single `numeric` value indicating how many dimensions to use or a vector of `numeric` values indicating which specific dimensions to use. This element is by default the number of columns in `$matrix` but can be modified through `boot.matrix()` or `dispRity()`.
 * `$call$bootstrap`: this is a `list` containing three elements:
    * `[[1]]`: the number of bootstrap replicates (`numeric`)
    * `[[2]]`: the bootstrap method (`character`)
    * `[[3]]`: the rarefaction levels (`numeric` vector)
 * `$call$disparity`: this is a `list` containing one element, `$metric`, that is a `list` containing the different functions passed to the `metric` argument in `dispRity`. These are `call` elements and get modified each time the `dispRity` function is used (the first element is the first metric(s), the second, the second metric(s), etc.).
